---
title: "Assignment 5: Data Visualization"
author: "Garrett Pearce"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single PDF file.
5. After Knitting, submit the completed exercise (PDF file) to the dropbox in Sakai. Add your last name into the file name (e.g., "Fay_A05_DataVisualization.Rmd") prior to submission.

The completed exercise is due on Monday, February 14 at 7:00 pm.



## Set up your session 
1. Set up your session. Verify your working directory and load the tidyverse 
and cowplot packages. Upload the NTL-LTER processed data files for nutrients 
and chemistry/physics for Peter and Paul Lakes (use the tidy 
[`NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv`] version) and 
the processed data file for the Niwot Ridge litter dataset (use the [`NEON_NIWO_Litter_mass_trap_Processed.csv`] version). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r}
#1 
getwd()
library(dplyr)
library(tidyverse)
library(cowplot)

peterpaul <- read.csv('./Data/Processed/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv')
niwotridge <- read.csv('./Data/Processed/NEON_NIWO_Litter_mass_trap_Processed.csv')

#2 
class(niwotridge$collectDate)
#Checking to see if the Niwot Ridge data column "collectDate" is read as a date- looks like it's not.

niwotridge$collectDate <- as.Date(niwotridge$collectDate, format = "%m/%d/%y")

peterpaul$sampledate <- as.Date(peterpaul$sampledate, format = "%m/%d/%y")

#Changed both date columns in the datasets to date format.

class(niwotridge$collectDate)
class(peterpaul$sampledate)

#Both look good!
```
## Define your theme
3. Build a theme and set it as your default theme.

```{r}
#3
newTheme <- theme(legend.position = "right", axis.text = element_text(color = "blue", size = 10), legend.title = element_text(size = 15, color = "red"))

#Setting a new (potentially ugly) theme and setting as default

theme_set(newTheme)
```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add a line of best fit and color it black. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and `ylim()`).

```{r}
#4
phos_Plot <- ggplot(peterpaul, aes(x = tp_ug, y = po4, color = lakename, shape = lakename)) + geom_point(alpha=0.7, size=2.5) + geom_smooth(color = 'black') + xlim(0,150) + ylim(0,50) + labs(title = 'Total Phosphorous by Phosphate') + xlab('Total Phosphorous') + ylab('Phosphate')

#  Note- I am not sure if geom_smooth() gives the best line of best fit here- ideally I would have liked straight lines, and to have one line of best fit for the graph as a whole, but these do give a solid overview of the trends for both lakes.
```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned.

```{r}
#5
temp_Plot <- ggplot(peterpaul, aes(x = month, y = temperature_C, color = lakename)) + geom_boxplot()
TP_Plot <- ggplot(peterpaul, aes(x = month, y = tp_ug, color = lakename)) + geom_boxplot()
TN_Plot <- ggplot(peterpaul, aes(x = month, y = tn_ug, color = lakename)) + geom_boxplot()
#Creating initial plots

all_Plots <- plot_grid(temp_Plot, TP_Plot, TN_Plot)
plot_Legend <- get_legend(all_Plots)
#Creating objects to align plots and show legend

final_Plot <- plot_grid(all_Plots, plot_Legend)
#Combining aligned plots and legend into final product
```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: 

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r}
#6

#7


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer:
